---
title: "Data Visualization in R"
author: "Prof. Harbert"
date: "Part 4"
output: 
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
  
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# [Home](https://devbioinformatics.org)


# Data Visualization

Today we will be using 'ggplot', an R package for data visualization. These tools will help you create publication quality plots and figures directly from your R code. If there is one thing that you get out of this class I hope that it is som confidence with these tools. 

What is 'ggplot'? The name refers to the "Grammar of Graphics". This refers to a structured language for building data visualizations. 

## ggplot resources

[Hadley Wickham's paper](http://vita.had.co.nz/papers/layered-grammar.pdf)

[The Grammar of Graphics Book](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448/ref=sr_1_1?keywords=grammar+of+graphics&qid=1568209903&s=gateway&sr=8-1)

[R for Data Science -- Data Visualization chapter](https://r4ds.had.co.nz/data-visualisation.html)

[ggplot documentation](https://ggplot2.tidyverse.org/reference/)



## Recent news: ggplot in the press

The BBC Visual and Data Journalism team uses R and ggplot to create their publication graphics. A few weeks ago they released their toolbox on [GitHub](https://github.com/bbc/bbplot). More from [Medium](https://medium.com/bbc-visual-and-data-journalism/how-the-bbc-visual-and-data-journalism-team-works-with-graphics-in-r-ed0b35693535)


# Visualization Practice

## Read Census Data

For this section we will read in a table of US Census Demographic data from https://www.kaggle.com/muonneutrino/us-census-demographic-data.


To read data into R we use:
read.table()
read.csv()
read.delim()

```{r, eval=TRUE}
census = read.csv('https://raw.githubusercontent.com/rsh249/applied_bioinformatics2020/master/data/census.csv')
head(census)

```

LOOK BACK: How do you write a table to a new file?


## ggplot2 Graphics

For help and an overview of all of the graphics types that are available in ggplot check out the [R ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)

```{r}
library(ggplot2)
```

After loading the ggplot2 library we have access to the visualization functions in the library. The general form of all ggplot calls will be:

```{R, eval=F}
# start a plot object
# declare the data you want to work with in ggplot
ggplot(data=somedata) +
    # define one or more 'geom'etric attributes.
    # These tell R what kind of graph we want (e.g., scatterplot, histogram)
    # define the aesthetic (first: what variables to use)
    geom_point(aes(x=xvar, y=yvar)) +
    # Edit things about how the graph looks
    theme()


```

### Scatterplots using 'geom_point()'

```{r}
ggplot(data = census) +
  geom_point(aes(x = Citizen, y = TotalPop))
```

## Graphs for single parameters:

### geom_histogram()

```{r}
ggplot(data=census) +
  geom_histogram(aes(x=Citizen), binwidth=100) 
```

But the x-axis is based on the maximum value, not the effective range so the distribution of the data is tough to read.

```{r}
ggplot(data=census) +
  geom_histogram(aes(x=Citizen), binwidth=100)+
  xlim(c(0,10000))
```

That also got rid of an artifact near zero. Let's inspect the data for 'Citizen' to see what data could be causing that:

```{r}
#summary(census$Citizen)
sum(census$Citizen <=0)
```

Are zero values real?

#### Sidetrack: Logical subsetting

We can use logical operators and the data.frame index notation to find ONLY those rows of data that have a zero value in the Citizen column.

```{r}
# test for Citizen ==0
logic = census$Citizen==0
# Then use this vector of logical results to index the data frame
zerocit= census[logic,] #make sure to use the rows position -> left of the column
#summary(zerocit)
```
### Density:

```{r}
ggplot(data=census) +
  geom_density(aes(x=Citizen)) +
  xlim(c(0,10000))
```

### For Discrete values: geom_bar()

```{r}
ggplot(data=census) +
  geom_bar(aes(x=State))

```
But, this is hard to read. A few fixes:

+ Rotate x-axis labels (Use 'theme()')
+ Sort states by count (Use library(forcats) -- Tools for working with categorical variables)

```{r}
library(forcats)
ggplot(data=census) +
  geom_bar(aes(x=fct_infreq(State))) +
  xlab('') +
  theme(axis.text.x  = element_text(angle=90))
```

### Challenge 1:

Pick a plot method appropriate for the data in the MeanCommute column. 

## Graphs for two parameters

### Continuous parameters

#### geom_point()

Scatterplots

```{R}
ggplot(data=census) + 
    geom_point(aes(x=IncomePerCap, y=Poverty))
```

See also: geom_jitter() for when you have overlapping points, and geom_label() to add text labels to points.

#### geom_smooth

Add a line. WARNING: The default is NOT a linear regression. 

```{r}
ggplot(data=census) + 
    geom_point(aes(x=IncomePerCap, y=Poverty), alpha=0.1) +
    geom_smooth(aes(x=IncomePerCap, y=Poverty)) # adds a line of best fit
```

To make this a plot with a linear model (Maybe Not the Best Idea) use the 'method="lm"' argument in geom_smooth:

```{R}
ggplot(data=census) + 
    geom_point(aes(x=IncomePerCap, y=Poverty), alpha=0.1) +
    geom_smooth(aes(x=IncomePerCap, y=Poverty), method='lm')
```


#### Countour and density plots

```{R}
ggplot(data=census) + 
    geom_hex(aes(x=IncomePerCap, y=Poverty)) 
```
```{R}
ggplot(data=census) + 
    geom_density2d(aes(x=IncomePerCap, y=Poverty)) 
```

### One Discrete X One Continuous parameter

#### Boxplots

```{R}
ggplot(data=census[1:10000,]) + 
    geom_boxplot(aes(x=State, y=Poverty)) 
```

#### Violin charts

```{R}
ggplot(data=census[1:10000,]) + 
    geom_violin(aes(x=State, y=Poverty)) 
```


### Challenge 2

Develop a plot showing the relationship between MeanCommute (time in minutes) and Transit (% using transit). 

## Changing the appearance of your plot:


### Color 

We can change colors of attributes. E.g., the points in a scatterplot
```{R}
#first make this smaller and look only for data from Mass
census2=census[census$State=='Massachusetts',]
ggplot(data=census2) + 
    geom_point(aes(x=IncomePerCap, y=Poverty, col='red'))
```

Or we can apply a range of colors that correspond to some categorical data for each row.

```{R}
ggplot(data=census2) + 
    geom_point(aes(x=IncomePerCap, y=Poverty, col=County), alpha=0.1)
```

### Themes

Every ggplot object has a theme attribute. We can manipulate those with the 'theme()' function, but there are ~1,000 moving pieces there. For now, here are some default theme functions that produce nice graphics:

#### theme_bw: black and white lines and background

```{R}
ggplot(data=census2) + 
    geom_point(aes(x=IncomePerCap, y=Poverty, col=County), alpha=0.1) +
  theme_bw()
```

#### theme_minimal: simplified

```{R}
ggplot(data=census2) + 
    geom_point(aes(x=IncomePerCap, y=Poverty, col=County), alpha=0.1) +
  theme_minimal()
```

#### theme_void: not usually what you want

```{R}
ggplot(data=census2) + 
    geom_point(aes(x=IncomePerCap, y=Poverty, col=County), alpha=0.1) +
  theme_void()
```

# Save your work

You should have been keeping your code in an R script file. Save that file now so that you can refer back to it later.

# Save your Plot

The function 'ggsave()' can save any ggplot object as an image. By default it will save the most recent plot in your session with:

```{R}
ggsave(file='lastgraph.png', height=6, width=6, units='in', dpi=600)
```

OR: Much better method --> Redirect your plot call to an object and ggsave that


```{R}
graph_obj <- ggplot(data=census2) + 
    geom_point(aes(x=IncomePerCap, y=Poverty, col=County), alpha=0.1) +
  theme_bw()

ggsave(graph_obj, file='income_vs_poverty.png', height=6, width=6, units='in', dpi=600)

```

# Homework:

Include notes on ggplot and code to create *at least one* of these plots in your blog for the week.




# [Home](https://devbioinformatics.org)
