---
title: "R Basics: Vectors, matrix objects, read files"
author: "Prof. Harbert"
date: "Part 2"
output: 
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
  
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goals

+ Practice more basic R commands
+ Write R code in the editor and run in the R console
+ Develop good Git habits
+ Read/write data files
+ Visualize with simple plots

# Basics

Open RStudio. 

Try the following in the R Console:

```{r, eval=F}
x=2
print(x) ##Print method
class(x)
```

What did that do? 'x=2' sets a variable named x equal to 2. Then we use the function print() to show us what is in the variable 'x'. Last, class() is a function that shows what kind of data is stored in the variable x.

*Recall*: We saw last session that R can perform basic math operations both with entered numbers or values stored in variables.

## Functions

In all programming languages most code is base on functions. R functions are always a single word or phrase with parentheses () at the end (e.g., print()). Arguments (options) and data can be passed to the function code.

We will see many functions going forward.

## Vectors

Often we need to create objects that store more than one value. A simple list of data is called a vector in R. To set up one:

```{r, eval=F}

x=seq(10:1) # A function to ceate a vector
class(x)
print(x)
```

R gives you control over what data in a vector you access if you are careful with what you are typing. 

The position in a vector is called the index and is represented by a number starting with 1. (*Note*: Some other programming languages start vectors at index '0', R is the odd language out)

```{R, eval=F}
print(x[1]) # First index of vector
print(x[1:5]) #Print all 5
print(x[c(1,3,5)]) #print only index 1, 3, and 5
```

## 2D Data

Most data, however, is not just a single list. Most of the time you will have rectangular data (think an Excel spreadsheet) where you have data stored in rows and columns.

A basic data structure for storing 2D data is the matrix object.

```{R, eval=F}

y = matrix(nrow=5, ncol=5) # create a 5x5 matrix
print(y) # see the matrix
class(y) # what is it?

y[1,1] = 5 # change a cell value
print(y) # see the change
```

Rows and columns. In 2D data structures there are index values for both rows and columns. Access individual cells using the name of the matrix with brackets [row number, column number]. 

```{R, eval=F}
y[1,1] # See value in row 1 column 1
y[,1] # see all of column 1

y[,1] = seq(1,5) # put data in column 1
y[,2] = seq(6,10)
y[,3] = seq(11,15)
y[,4] = seq(16,20)
y[,5] = seq(11,15)

print(y)

```


# Read table/tab/csv/txt text files:
read.table()
read.csv()
read.delim()

```{r}
file = 'https://raw.githubusercontent.com/rsh249/applied_bioinformatics2020/master/data/mtcars.csv' #URL to text file with data

cars = read.table(file, header=T, sep = ',') # Read a comma separated values file
head(cars)

cars = read.csv(file)

cars = read.delim(file, sep=',')


```

# Write Files

```{r}
write.table(cars, file='mtcars.tab')
write.csv(cars, file='mtcars.csv')
```

Where are these files now?

# Next: Visualize data


# [Home](https://devbioinformatics.org)









