---
title: "R Basics: Vectors, matrix objects, read files"
author: "Prof. Harbert"
date: "Part 3"
output: 
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Recap

In the last session we saw the RStudio interface and started typing at the R command prompt to practice basic R calculations and variable setting. Then we saw how to read files into R.

# More on Files

As a review: Here is how we were reading our data files last time:

```{r}
file = 'https://raw.githubusercontent.com/rsh249/applied_bioinformatics2020/master/data/mtcars.csv' #URL to text file with data

cars = read.table(file, header=T, sep = ',') # Read a comma separated values file
head(cars)
```

## Write Files

```{r}
write.table(cars, file='mtcars.tab')
write.csv(cars, file='mtcars.csv')
```

Where are these files now?

## A note on Paths

Ask R where you are:

```{R}
getwd() # Will print your current path
```

To use data stored on a computer we need to tell R where it is. This is done using paths. 

Paths can be *absolute*:

```{r}
#cars <- read.csv('/home/rharbert/GitHub/bioinformatics/docs/mtcars.csv') #Linux path type for server
#cars <- read.csv('/Users/rharbert/.../mtcars.csv') #Windows type if working on a personal machine
```
**Your paths will be different**

Folders/Directories are separate by / with the file name at the end.
    
Paths can also be relative:

```{r}
cars <- read.csv('mtcars.csv')
```

"From where I am open the mtcars file "

## Your are Here

How do you figure out what the path *should* be?

```{r}
getwd()
```

Prints the current path of your R session.

If you need to change the directory you are working in, use:

```{R, eval=F}
setwd('/home/rharbert/GitHub/bioinformatics/docs/')
```

OR the File pane in RStudio.


# Basic plotting

One of R's biggest advantages is the ability to create high quality graphics in nearly any format or style. Today we will be working with the basic plotting features of R. These are good, but limited at times. 

Later we will take a look at the [ggplot library](https://ggplot2.tidyverse.org/). ggplot is the current 'state of the art' in graphics for R. 

```{r}
cars=read.csv('mtcars.csv')
```


```{r}
head(cars)
plot(cars)
```

OK. That was not so great. Let's try somethnig more useful for visualizing these data. We can tell plot() which columns we want to create a scatterplot for:

```{r}
colnames(cars)
plot(cars[,'cyl'], cars[,'mpg'])
```

OR we can create other types of plots by calling other functions. e.g., a histogram of boxplot:

```{r}
hist(cars[,'mpg'])

```


```{r}
boxplot(cars[,'hp'])
```

# Homework:

1) *DISCUSSION*: In the slack [#discussion](https://bio332applied-4mi2242.slack.com/archives/CS8TRGE0K) channel we will host this week's discussion. Please search online for an R learning resource for R beginners. This could be a book/eBook, a web tutorial, youtube video, or something else. Post what you find. Look through the resource you found and tell the group about it (what it is, how it is structured, who it is targetted to).
2) *BLOG*: Set up your Wordpress account @ https://devbioinformatics.org/bio332. Then create a blog post where you detail at least three concepts in the R coding we covered today. You need to include code examples. To add code chunks look for the code formatting options when you add a new content section to your post (See guide: https://en.support.wordpress.com/wordpress-editor/blocks/code-block/). Discuss what the code is doing and what the resulting process returns. If you are unclear on how something works look online (see reference search from HW#1) and then ask [#questions](https://bio332applied-4mi2242.slack.com/archives/CSRS46M7H) on Slack.







# [Home](https://devbioinformatics.org)



